---
title: "MDA cell line normalization"
output:
  html_document:
    toc: TRUE
bibliography : A2.bib
---
```{r}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE,echo = TRUE, results = "hide", tidy = TRUE)
```



# A1 : Data Exploration, normalization, and mapping

## Set up 
```{r,setup, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
if (!requireNamespace("BiocManager",quietly = TRUE)){
  install.packages("BiocManager")}
library(BiocManager)
if (!requireNamespace("GEOmetadb",quietly = TRUE)){
  BiocManager::install('GEOmetadb')}
if (!requireNamespace("GEOquery",quietly = TRUE)){
  BiocManager::install('GEOquery')}
if (!requireNamespace("edgeR",quietly = TRUE)){
  BiocManager::install("edgeR")}
if (!requireNamespace("biomaRt",quietly = TRUE)){
  BiocManager::install("biomaRt")}
if (!requireNamespace("DT",quietly = TRUE)){
  install.packages('DT')}
library(knitr)
library(edgeR)
library(tidyverse)
library(GEOmetadb)
library(biomaRt)
library(DT)
```

Package references used in A1
[@biocmanager;@geometadb;@edger;@knitr;@tidyverse;@biomart;@geoquery;@DT;@lecture4;@lecture5]


```{r,echo=FALSE,results='hide',message=FALSE,warning=FALSE}
# GEO description of my dataset
# GSE Series Matrix only specific for microarray 

GSE152201<-getGEO("GSE152201", GSEMatrix=FALSE)
#Meta() get meta data
DT::datatable(data.frame(head(Meta(GSE152201))), options = list(scrollX=T))
GPLList(GSE152201)
head(Meta(GSE152201))
# get current flatform 
current_platform=names(GPLList(GSE152201))[1]
# information about the platform
current_platform_infor = Meta(getGEO(current_platform))
names=names(current_platform_infor)
names
```

## Basic information about my dataset 

**Platform title**:`r current_platform_infor$title`

**Submission data**:`r current_platform_infor$submission_date`

**Last update data**:`r current_platform_infor$last_update_date`

**Organism**:`r current_platform_infor$last_update_date`

**Number of GEO datasets that use this technology**:`r length(current_platform_infor$series_id)`

**Number of GEO samples that use this technology**: `r length(current_platform_infor$sample_id)`

```{r}
# get supplementary file 
seq_files =getGEOSuppFiles("GSE152201")
file_name =rownames(seq_files)
file_name
exp_data = read.delim(file_name[1],header=TRUE,check.names = FALSE)
head(exp_data)
```



```{r,echo=FALSE,message=FALSE}
#look at the data first 
colnames(exp_data)[1]<-"ensembl_id"
DT::datatable(data.frame(exp_data[1:5,c(1,2,5,8,11)]), options = list(scrollX=T))
# the first column is not what I want, break into two column
exp_data<-exp_data %>% separate(ensembl_id,c("gene_name","gene_id"),sep = "\\_",extra = "merge", fill = "right")
exp_data<-exp_data %>% separate(gene_id,c("gene_id"),extra = "drop")
```



## Sample Data table

My dataset contains data of two cell lines, I am using the data from the MDA cell line
```{r, results='hold'}
# My dataset contains data of two cell lines, I am using the data from the MDA cell line
DT::datatable(data.frame(exp_data[1:5,c(1,2,9:14)]), options = list(scrollX=T),caption = "Table 1 :My data sample")

```
[@DT]

```{r,echo=FALSE}
# define the group
samples <- data.frame(lapply(colnames(exp_data)[9:14],FUN=function(x){unlist(strsplit(x, split = "\\_"))[c(1,2,4)]})) 
colnames(samples) <- colnames(exp_data)[9:14] 
rownames(samples) <- c("cell_lines","treatment","replicates") 
samples <- data.frame(t(samples))
knitr::kable(samples[1:6,1:3],format = "html")
```

## Filter low count

```{r, results='hide'}
# select the mda cell line data
exp_data_mda <-exp_data[,c(1,2,9:14)]
# converts it to count per million
count_per_million_mda=cpm(exp_data[,3:8])
# change the row name
rownames(count_per_million_mda)<-exp_data_mda[,"gene_id"]
# keep cpm greater than 3 per row
keep = rowSums(count_per_million_mda>1 ) >= 3
exp_data_filtered =exp_data_mda[keep,]
summarised_gene_count <-sort(table(exp_data_filtered$gene_id),decreasing = TRUE)
# sort it in decreasing order 
summarised_gene_count_unfiltered <-sort(table(exp_data_mda$gene_id),decreasing = TRUE)
# show an example

knitr::kable(summarised_gene_count[which(summarised_gene_count>1)][1:6],format = "html")
dim(exp_data_mda)
#how many genes are keptn
low_count_filtered<-nrow(exp_data_filtered)
```

## Visulize distribution after Normalization

```{r}
# filter outlier using normalization 
filtered_data_matrix <- as.matrix(exp_data_filtered[,3:8]) 
# change the row names
rownames(filtered_data_matrix) <- exp_data_filtered$gene_id
# create a DGEList for normalization
d = DGEList(counts=filtered_data_matrix, group=samples$treatment)
#calculate normalization factor
d_TMM = calcNormFactors(d)
normalized_counts <- cpm(d_TMM)
#number of outlier filtered
nrow(exp_data_filtered)-nrow(normalized_counts)
```


```{r,echo=FALSE,message=FALSE,results="hide"}
#normalized plot
log_cpm_data<-log2(cpm(exp_data_filtered[,3:8]))
data_density<-apply(log2(normalized_counts),2,density)
data_density
#set the x and y limit of the plot
xlim<-0;ylim<-0
for(i in 1:length(data_density)){
  xlim<-range(c(xlim,data_density[[i]]$x));
  ylim<-range(c(ylim,data_density[[i]]$y))
}
#color
cols<-rainbow(length(data_density))
#lines
ltys<-rep(1,length(data_density))
#plot
plot(data_density[[1]],xlim=xlim,ylim=ylim,type="n",
     ylab="Smoothing density",main="Figure 1: Normalized Distribution", sub = "MDA cell line" ,xlab="normalized counts in log2-cpm", cex.lab=0.8)
for (i in 1:length(data_density)) {lines(data_density[[i]],col=cols[[i]],lty=ltys)
}
#create legend
legend("topright", colnames(log_cpm_data),col=cols, lty=ltys, 
       cex=0.6, border ="blue",text.col = "green4",merge = TRUE, 
       bg = "gray90")
```
[@edger]

```{r,echo=FALSE,include=FALSE}
# There is another PCA plot in A2 so I did not include this
plotMDS(d_TMM,labels = rownames(samples),
        col=sample(c("blue","red"))[factor(samples$treatment)])
legend("topright",c("ETOH" ,"DEX"),pch = 19,fill = c("blue", "red"))
title(main = " Principle Component Plot", sub = "MDA cell line")

```

## Dispersion

Small variation as indicated by the graph 

```{r,echo=FALSE}
#dispersion: how much your variance deviates for the mean 
#low count high variation
model<-model.matrix(~samples$treatment+0)# 
dispersion<-estimateDisp(d_TMM,model)
plotBCV(dispersion,col.tagwise = "blue",col.common = "red")
title(main = "Figure 2:  Dispersion Plot", sub = "MDA cell line")
```
[@edger]

## Identifier mapping

```{r,echo=FALSE,results='hide'}
#get the right dataset
listMarts()
# different version maybe better mapping for data at that year
listEnsemblArchives()
#connect to the desired mart
ensembl<-useMart("ensembl")
#get the human dataset
datasets<-listDatasets(ensembl)
kable(head(datasets[grep(datasets$dataset,pattern ="sapiens"),]),format = "html")
ensembl=useDataset("hsapiens_gene_ensembl",mart = ensembl)
```

```{r,echo=FALSE,results='hide'}
#make sure your dataset's gene name is correct 
#we want to map our ensemble id to HGNC symbol
#we are going to get both ensemble id and HGNC symbol back in order to join the table 
#filter
dim(listFilters(ensembl))
all_biomart_human_filter<-listFilters(ensembl)
kable(all_biomart_human_filter[grep(all_biomart_human_filter$name,pattern="ensembl"),],format = "html")
# ENST is transcript, ENSG is gene, ENSP is protein
#search for a given attribute, in this case HGNC id 
kable(searchAttributes(mart=ensembl,"hgnc"),format="html")
```


```{r}
# map my ensembl id to hgnc symbol using the correct mart and biomart filter,  
conversion_mda<-"mda_cell_line.rds"
if(file.exists(conversion_mda)){
  mda_cell_line<-readRDS(conversion_mda)
}else{
  mda_cell_line<-getBM(attributes = c("ensembl_gene_id","hgnc_symbol"),
                    filters = "ensembl_gene_id",
                    value = exp_data_filtered$gene_id,
                    mart = ensembl)
  saveRDS(mda_cell_line,conversion_mda)
}
```
[@biomart]

```{r,echo=FALSE, results="hold"}
length(which (rownames(normalized_counts)%in%
         mda_cell_line$ensembl_gene_id))
nrow(normalized_counts)
not_mapped = length(which (rownames(normalized_counts)%in%
         mda_cell_line$ensembl_gene_id)) - nrow(normalized_counts)
```


```{r,echo=TRUE}
#merge to include rest of the data
normalized_counts_merge <- merge(mda_cell_line,normalized_counts,
                                 by.x = 1, by.y = 0, all.y=TRUE)
#how many are missing
kable(normalized_counts_merge[1:5,1:5],type = "html",caption = "Table 3 : missing hgnc symbols")

ensembl_id_missing_gene <- normalized_counts_merge$ensembl_gene_id[
  which(is.na(normalized_counts_merge$hgnc_symbol))]#1207
no_symbol = length(ensembl_id_missing_gene) 
# old hgnc from my original dataset
old_mapping <- merge(exp_data_filtered[,1:2],data.frame(ensembl_id_missing_gene),by.x = 2,by.y = 1)
# some of the old hgnc symbol
kable(old_mapping[1:10,], type="html",caption = " Table 4: unmapped ensembl id  found in original dataset")
# how many ribosomal protein there are
count = nrow(old_mapping[grep(old_mapping$gene_name, pattern = "^RP"),])
```

I merge the data set and `r no_symbol` observations are missing, `r count`are ribosomal genes but it is not really important for this dataset. The coverage now around 95%

```{r, echo=TRUE}
# missing hgnc symbol 
missing_ids_subset <- normalized_counts_merge[   which(is.na(normalized_counts_merge$hgnc_symbol)),]
# merge the missing symbol with old mapping
missing_ids_subset_withids <- merge(old_mapping,missing_ids_subset,by.x=1,by.y=1)
# filter out the NAs caused by the merge 
missing_ids_subset_withids <- missing_ids_subset_withids[-3]
# set the same column names
colnames(missing_ids_subset_withids)[1:2] <- colnames(normalized_counts_merge)
# bind the old mapping with the rows that have missing hgnc symbol
finalized_normalized_counts<-rbind(normalized_counts_merge[which(!is.na(normalized_counts_merge$hgnc_symbol)),],missing_ids_subset_withids)
```

```{r}
# drop some NAs
finalized_normalized_counts <- normalized_counts_merge %>% drop_na(.)
# check and filter duplicated ensembl id
length(unique(finalized_normalized_counts$ensembl_gene_id))
length(unique(finalized_normalized_counts$hgnc_symbol))
finalized_normalized_counts <- distinct(finalized_normalized_counts, ensembl_gene_id, .keep_all = TRUE) 
any(is.na(finalized_normalized_counts$hgnc_symbol))


#write.table(finalized_normalized_counts, file =  file.path(getwd(),"Data",                 "GSE152201_finalized_normalized_counts_mda_2022.txt"), sep = "\t")
```


